Intrinsic alignments from projected tidal fields: non-linear impact on cosmic shear probes

### 2024-06-21 sections/2_cosmic_shear_stats.tex

Shear two-point correlation functions (\gtwopcf hereafter) can be predicted with percent-level precision and are therefore an ideal quantity to validate weak lensing simulations.
In the Limber approximation\footnote{See \citet{Kilbinger17} for a comparison between the Limber approximation and the exact calculations.}, the tomographic lensing power spectrum $C^{ij}_{\ell}$, obtained for combinations of redshift bins $i$ and $j$, is calculated from an integral over the three-dimensional matter power spectrum $P_{\delta}(k, z)$ as:
  \begin{eqnarray}
C_{\ell}^{ij} = \int_0^{\chi_{\rm H}}  \frac{q^i(\chi) \,q^j(\chi)}{\chi^2} \, P_{\delta}\, \bigg(\frac{\ell+1/2}{\chi},z(\chi)\bigg) \ {\rm d}\chi,
\label{eq:C_ell}
\end{eqnarray}
where $\ell$ are angular multipoles, $k$ are amplitudes of Fourier modes, $\chi_{\rm H}$ is the comoving distance to the horizon, $c$ is the speed of light, $H_0$ the Hubble parameter, and the lensing kernels $q^{i}$ and $q^j$ are given by:
\begin{eqnarray}
q^{i}(\chi) = \frac{3}{2}\Omega_{\rm m} \, \bigg(\frac{H_0}{c} \bigg)^2 \frac{\chi}{a(\chi)} \int_{\chi}^{\chi_{\rm H}} n^i(\chi')\frac{{\rm d}z}{{\rm d} \chi'}\frac{\chi' - \chi}{\chi'}{\rm d}\chi' \, .
\label{eq:q_lensing}
\end{eqnarray}
In the previous expression, $n^i(z)$ refers to the redshift distribution  in tomographic bins `$i$', while $a(\chi)$ is the scale factor at comoving distance $\chi$ from the observer.
The matter power spectrum is computed from  {\sc Halofit} \citep{Takahashi2012} in this work, however other public tools provide accurate predictions, including \eg  {\sc HMcode} \citep{HMCode2020}, the {\sc EuclidEmulator} \citep{EuclidEmulator}, the {\sc Bacco} emulator \citep{BACCOEmulator}, or the {\sc MiraTitan} emulator \citep{miraTitan}. 
Predictions for the \gtwopcf are finally computed from Eq. (\ref{eq:C_ell}) as:
\begin{eqnarray}
\xi_{\pm}^{ij}(\vartheta) = \frac{1}{2\pi} \int_0^{\infty} C_{\ell}^{ij} \, J_{0/4}(\ell \vartheta) \, \ell \, {\rm d}\ell,
\label{eq:xipm_th}
\end{eqnarray}
where $J_{0/4}(x)$ are Bessel functions of the first kind. 
In this paper, these calculations are carried out by the public\footnote{{\sc CosmoSIS}:https://cosmosis.readthedocs.io/en/latest/ } {\sc CosmoSIS} cosmological inference package \citep{cosmoSIS}.
The redshift distribution $n (z)$ is taken from the LSST Year-1 forecast \citep{LSST-SRD}:
\begin{eqnarray}
n (z) =  z^2 \: {\rm exp} \left[-\left(\frac{z}{z_0}\right)^\alpha \right]\, ,
\label{eq:nz}
\end{eqnarray}
with pivot redshift $z_0=0.13$ and a power-law index $\alpha = 0.78$.
The distribution is normalised to provide a number density of $n_{\rm eff} = 3.0 $  galaxies per arcmin$^{2}$.
This lower than the expected number density for the first data release ($n_{\rm eff}\sim10$ galaxies arcmin$^{-2}$), but is large enough to validate our methods, and makes the calculation more tractable.
This global $n(z)$ is further split into five equi-populated tomographic bins, and smoothed with a Gaussian filter of width $\sigma = 0.05 \: (1 + z)$ to mimic the photometric selection process, shown by the solid lines in Fig. \ref{fig:Nz}.

@@

In addition to the probes mentioned above, we further consider the following convergence map-based statistics:
\begin{enumerate}
    \item \textit{Convergence three-point correlation function}: \JHD{Alejandro, can you summarise the measurements and modelling, and provide references?}
    \item \textit{Peaks and Minima}: Local maxima and minima in the {\sc Healpix} convergence maps are counted in bins of signal-to-noise ratios, $\nu$, computed from the global  noise levels in the survey. Specifically, pure noise maps $\mathcal{N}$ are constructed from generating pure Gaussian maps with mean set to zero and variance set to $\sigma^2 =  \sigma_\epsilon^2 / (2 \Delta \Omega_{\rm pix} n_{\rm gal})$. Here, $\Delta \Omega_{\rm pix}$ is the average pixel area, $n_{\rm gal}$ is the mean galaxy density in our survey (0.6 gal / arcmin$^2$ per tomographic bin)  and $\sigma_\epsilon$ is the intrinsic dispersion in galaxy shapes, per component, which we set to 0.27.
   Peak count statistics are amongst the most widely used non-Gaussian lensing statistics, as it is simple to measure yet highly sensitive to non-Gaussian structures.
    Although some theoretical models have been developed to describe the largest peaks \citep[see][]{Shan18, HSCY1_Peaks_th}, we report here the results on a much wider range of statistics, which can only be accurately modelled from numerical simulations themselves.
    \item \textit{Lensing PDF}:  \JHD{Cora/Lina, can you summarise the measurements and modelling, and provide references?}
   % \item  \textit{Minkowski functionals}: \JHD{Nisha, can you summarise the measurements and modelling, and provide references?}
\end{enumerate}

>>>>

Shear two-point correlation functions (\gtwopcf hereafter) can be predicted with percent-level precision and are therefore an ideal quantity to validate weak lensing simulations.
In the Limber approximation\footnote{See \citet{Kilbinger17} for a comparison between the Limber approximation and the exact calculations.}, the tomographic lensing power spectrum $C^{ij}_{\ell}$, obtained for combinations of redshift bins $i$ and $j$, is calculated from an integral over the three-dimensional matter power spectrum $P_{\delta}(k, z)$ as:
  \begin{eqnarray}
C_{\ell}^{ij} = \int_0^{\chi_{\rm H}}  \frac{q^i(\chi) \,q^j(\chi)}{\chi^2} \, P_{\delta}\, \bigg(\frac{\ell+1/2}{\chi},z(\chi)\bigg) \ {\rm d}\chi,
\label{eq:C_ell}
\end{eqnarray}
where $\ell$ are angular multipoles, $k$ are amplitudes of Fourier modes, $\chi_{\rm H}$ is the comoving distance to the horizon, $c$ is the speed of light, $H_0$ the Hubble parameter, and the lensing kernels $q^{i}$ and $q^j$ are given by:
\begin{eqnarray}
q^{i}(\chi) = \frac{3}{2}\Omega_{\rm m} \, \bigg(\frac{H_0}{c} \bigg)^2 \frac{\chi}{a(\chi)} \int_{\chi}^{\chi_{\rm H}} n^i(\chi')\frac{{\rm d}z}{{\rm d} \chi'}\frac{\chi' - \chi}{\chi'}{\rm d}\chi' \, .
\label{eq:q_lensing}
\end{eqnarray}
In the previous expression, $n^i(z)$ refers to the redshift distribution  in tomographic bins `$i$', while $a(\chi)$ is the scale factor at comoving distance $\chi$ from the observer.
The matter power spectrum is computed from  {\sc Halofit} \citep{Takahashi2012} in this work, however other public tools provide accurate predictions, including \eg  {\sc HMcode} \citep{HMCode2020}, the {\sc EuclidEmulator} \citep{EuclidEmulator}, the {\sc Bacco} emulator \citep{BACCOEmulator}, or the {\sc MiraTitan} emulator \citep{miraTitan}. 
Predictions for the \gtwopcf are finally computed from Eq. (\ref{eq:C_ell}) as:
\begin{eqnarray}
\xi_{\pm}^{ij}(\vartheta) = \frac{1}{2\pi} \int_0^{\infty} C_{\ell}^{ij} \, J_{0/4}(\ell \vartheta) \, \ell \, {\rm d}\ell,
\label{eq:xipm_th}
\end{eqnarray}
where $J_{0/4}(x)$ are Bessel functions of the first kind. 
In this paper, these calculations are carried out by the public\footnote{{\sc CosmoSIS}:https://cosmosis.readthedocs.io/en/latest/ } {\sc CosmoSIS} cosmological inference package \citep{cosmoSIS}.
The redshift distribution $n (z)$ is taken from the LSST Year-1 forecast \citep{LSST-SRD}:
\begin{eqnarray}
n (z) =  z^2 \: {\rm exp} \left[-\left(\frac{z}{z_0}\right)^\alpha \right]\, ,
\label{eq:nz}
\end{eqnarray}
with pivot redshift $z_0=0.13$ and a power-law index $\alpha = 0.78$.
The distribution is normalised to provide a number density of $n_{\rm eff} = 3.0 $  galaxies per arcmin$^{2}$.
This lower than the expected number density for the first data release ($n_{\rm eff}\sim10$ galaxies arcmin$^{-2}$), but is large enough to validate our methods, and makes the calculation more tractable.
This global $n(z)$ is further split into five equi-populated tomographic bins, and smoothed with a Gaussian filter of width $\sigma = 0.05 \: (1 + z)$ to mimic the photometric selection process, shown by the solid lines in Fig. \ref{fig:Nz}\footnote{The LSST SRD-Y1 forecasted redshift distribution can be generated here: github.com/LSSTDESC/CCLX/blob/master/srd\_redshift\_distributions.py}.

@@

In addition to the probes mentioned above, we further consider the following convergence map-based statistics:
\begin{enumerate}
    \item \textit{Convergence three-point correlation function}: \JHD{Alejandro, can you summarise the measurements and modelling, and provide references?}
    \item \textit{Peaks and Minima}: Local maxima and minima in the {\sc Healpix} convergence maps are counted in bins of signal-to-noise ratios, $\nu$, computed from the global  noise levels in the survey. Specifically, pure noise maps $\mathcal{N}$ are constructed from generating pure Gaussian maps with mean set to zero and variance set to $\sigma^2 =  \sigma_\epsilon^2 / (2 \Delta \Omega_{\rm pix} n_{\rm gal})$. Here, $\Delta \Omega_{\rm pix}$ is the average pixel area, $n_{\rm gal}$ is the mean galaxy density in our survey (0.6 gal / arcmin$^2$ per tomographic bin)  and $\sigma_\epsilon$ is the intrinsic dispersion in galaxy shapes, per component, which we set to 0.27.
   Peak count statistics are amongst the most widely used non-Gaussian lensing statistics, as it is simple to measure yet highly sensitive to non-Gaussian structures.
    Although some theoretical models have been developed to describe the largest peaks \citep[see][]{Shan18, HSCY1_Peaks_th}, we report here the results on a much wider range of statistics, which can only be accurately modelled from numerical simulations themselves.
    \item \textit{Lensing PDF}:  \JHD{Cora/Lina, can you summarise the measurements and modelling, and provide references? Cite \citet{LensingPDF}}
   % \item  \textit{Minkowski functionals}: \JHD{Nisha, can you summarise the measurements and modelling, and provide references?}
\end{enumerate}

### 2024-06-21 sections/4_simulations.tex

Ray-tracing is performed in the Born approximation, summing over the mass shells using a $\chi$-integral similar to Eq. (\ref{eq:C_ell}), but with only one power of $(q^i/\chi)$ instead of two, and replacing $P_\delta$ with $\delta_i(\theta,\phi)$.
A source plane is placed at the high-redshift edge of every mass planes, resulting in a series of convergence  maps $\kappa_i(\theta,\phi)$ that are subsequently transformed into shear maps $\gamma_{1/2, i}(\theta,\phi)$ using Eq. (\ref{eq:KS}).
We finally position galaxies in the light cone following three distinct algorithm, each impacting the strength of the IA signal:
\begin{enumerate}
\item {\it Random:} galaxies are distributed randomly on the octant, thereby reproducing one of the  fundamental assumptions in the NLA and TT models. 
\item {\it Linear bias:} galaxy positions are sampled from the mass sheets smoothed\footnote{We also tried sampling the field with $0.1$ and $0.5$ $h^{-1}\mathrm{Mpc}$ but this resulted in noisier results.} with a 1.0 $h^{-1}\mathrm{Mpc}$ (comoving) beam, assuming a linear bias of $b_{\rm TA}$, thereby building the one of the key assumptions of the \dNLA and \dTT models. 
We assume $b_{\rm TA}=1.0$ as our fiducial case but also consider $b_{\rm TA}=2.0$ to test  the  model flexibility. 
\item{\it Non-linear bias:} galaxy positions are obtained by populating dark matter haloes with the Halo Occupation Distribution prescription described in \citet{cosmoDC2}. 
\end{enumerate}
Note that while cases (i) and (ii) exactly reproduce the SRD-Y1 $N(z)$ presented in Fig. \ref{fig:Nz} (solid lines), the $N(z)$ in the latter case is not a perfect match, as shown with the dashed lines, although the agreement is good enough for our goal; we return to this in Sec. \ref{subsec:HOD}, where we describe the HOD model and related quantities.

@@

\begin{figure*}
\caption{Density field and the associated projected tidal field tensors.}
\label{fig:maps}
\end{figure*}

@@

%We employ a numerical technique worth mentioning here: the Fourier transforms involved in computing Eq. (\ref{eq:sij_2D}) are computed from the full (projected) periodic boxes of the simulations, then interpolated on the light-cones. We find that tidal field computed directly from the $10\times10$ deg$^2$ light-cones introduces significant large scales features in the tidal field maps, largely caused by  the non-periodic boundary conditions, which our method avoids.
Projected tidal field maps $s_{11}$, $s_{22}$, and $s_{12}$ are constructed that way for each mass sheets; Fig. \ref{fig:maps} shows the  three tidal fields and the mass maps for one of them.
We can clearly see the connection between all maps around over-dense regions. {\it (Include a figure of $s_{ij}$ and $\delta$)}.

>>>>

Ray-tracing is performed in the Born approximation, summing over the mass shells using a $\chi$-integral similar to Eq. (\ref{eq:C_ell}), but with only one power of $(q^i/\chi)$ instead of two, and replacing $P_\delta$ with $\delta_i(\theta,\phi)$.
A source plane is placed at the high-redshift edge of every mass planes, resulting in a series of convergence  maps $\kappa_i(\theta,\phi)$ that are subsequently transformed into shear maps $\gamma_{1/2, i}(\theta,\phi)$ using Eq. (\ref{eq:KS}).
We finally position galaxies in the light cone following three distinct algorithm, each impacting the strength of the IA signal:
\begin{enumerate}
\item {\it Random:} galaxies are distributed randomly on the octant, thereby reproducing one of the  fundamental assumptions in the NLA and TT models. 
\item {\it Linear bias:} galaxy positions are sampled from the mass sheets smoothed\footnote{We also tried sampling the field with $0.1$ and $0.5$ $h^{-1}\mathrm{Mpc}$ but this resulted in noisier results.} with a 1.0 $h^{-1}\mathrm{Mpc}$ (comoving) beam, assuming a linear bias of $b_{\rm TA}$, thereby building the one of the key assumptions of the \dNLA and \dTT models. 
We assume $b_{\rm TA}=1.0$ as our fiducial case but also consider $b_{\rm TA}=2.0$ to test  the  model flexibility. 
\item{\it Non-linear bias:} galaxy positions are obtained by populating dark matter haloes with the Halo Occupation Distribution prescription described in \citet{cosmoDC2} and used recently in \citet{TXPipe}\footnote{We used version: {\tt skysim5000\_v1.2}.}. Specifically, we select objects over a patch of 732.20 deg$^2$ given by  0$<$RA$<$ 20 and -36.61$<$DEC$<$0, we apply a magnitude cut of mag$_r<$24.8, and further downsample randomly to retain 20\% of the galaxies, approaching the global galaxy redshift distribution and number density used with the other simulation catalogues (Eq. \ref{eq:nz}).
The sample is further split into 5 tomographic bins by calculating the probability a given galaxy in the HOD sample lies in each bin, based on its true redshift. This is simply done by interpolating each of the tomographic $N(z)$ shown in Fig. 1 at the galaxy?s true redshift, then using these as weights to assign a tomographic bin in a random draw. After a galaxy is assigned, it is removed from the sample to avoid double counting. This method allows us to capture some of the effect of photo-z uncertainty, as it gives all galaxies a small but non-zero probability of being included in tomographic bins outside of their true redshifts. The resulting in the $n(z)$ presented in Fig. \ref{fig:Nz} with the dashed lines, closely matching the target $n(z)$. Compared to the analytical SRD-Y1, the mean redshifts of the five tomographic bins are shifted by [-0.027, -0.018,  0.003, -0.010,  0.038], respectively, which we take into account when making predictions for these HOD lensing mocks.
\end{enumerate}
%Note that while cases (i) and (ii) exactly reproduce the SRD-Y1 $N(z)$ presented in Fig. \ref{fig:Nz} (solid lines), the $N(z)$ in the latter case is not a perfect match, as shown with the dashed lines, although the agreement is good enough for our goal; we return to this in Sec. \ref{subsec:HOD}, where we describe the HOD model and related quantities.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{graphs/TidalFieldOuterRim_V0}
\caption{Density field (top)  and the associated projected tidal field tensors $s_{11}$, $s_{22}$ and $s_{12}$, from left to right ({\it Improve this figure...}).}
\label{fig:maps}
\end{figure}

@@

%We employ a numerical technique worth mentioning here: the Fourier transforms involved in computing Eq. (\ref{eq:sij_2D}) are computed from the full (projected) periodic boxes of the simulations, then interpolated on the light-cones. We find that tidal field computed directly from the $10\times10$ deg$^2$ light-cones introduces significant large scales features in the tidal field maps, largely caused by  the non-periodic boundary conditions, which our method avoids.
Projected tidal field maps $s_{11}$, $s_{22}$, and $s_{12}$ are constructed that way for each mass sheets; Fig. \ref{fig:maps} shows the  three tidal fields and the underlying density maps for one of them.
We can clearly see the connection between all maps around over-dense regions. .
